---
output: html_document
editor_options: 
chunk_output_type: console
---
EDA on Prosper Loan Data by Dilip Rajkumar
========================================================


```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.

suppressMessages(library(ggplot2))
suppressMessages(library(ggthemes))
suppressMessages(library(dplyr))
suppressMessages(library(knitr))
suppressMessages(library(gridExtra))
suppressMessages(library(RColorBrewer))
suppressMessages(library(corrplot))
library(lubridate)
library(tidyr)
library(maps)
library(mapproj)
library(plotly)
library(GGally)
```

```{r echo=FALSE, Load_the_Data}
setwd("D:/OneDrive/Documents/Udacity - Data Analyst Nanodegree - Term 2/eda-course-materials/FinalProject")
# Load the Data
pld <- read.csv('prosperLoanData.csv')
```

```{r echo = FALSE, global_options}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

> **Abstract**: [Prosper](https://www.prosper.com/) is America's first peer-to-peer
lending marketplace, with more than 2 million members and over $2,000,000,000 in
funded loans. Here we use the data available to the public (last updated on 
March 11th, 2014) from [Udacity](https://docs.google.com/document/d/1qEcwltBMlRYZT-l699-71TzInWfk4W9q5rTCSvDVMpc/pub?embedded=true), which contains all the listings and loans ever created 
with 81 variables on each loan/listing, to do some data analysis. 

#### Structure of the Dataset
What are the dimensions of the dataset?
```{r echo=FALSE}
dim(pld)
```

Let's take a look at the sheer size of this dataset:
```{r}
str(pld)
```
This data set contains 113,937 loans with 81 variables such as loan amount, 
borrower rate (or interest rate), current loan status, borrower income, borrower
employment status, borrower credit history, Prosper score etc.. and the latest 
payment information.

#### Exploring and Cleaning Variables

The variable _'ListingCategory'_ shows the reason a loan was taken, such as for 
debt consolidation or maybe a student loan. Unfortunately, the data is stored 
numerically, with the key being listed [here](https://docs.google.com/spreadsheets/d/1gDyi_L4UvIrLTEC6Wri5nbaMmkGmLQBk-Yx3z0XDEtI/edit#gid=0).
  It makes much more sense to see the listing as what they are, rather than the 
arbitrary number it's been given. For that, I'll have to change the 
_'ListingCategory'_ variable from numeric to a factor:

<br>

```{r, echo = TRUE,message=FALSE, warning=FALSE}
# Convert ListingCategory from numeric to factor variable using the keys given
# in the Google Spreadsheet.
labels <- c("Debt Consolidation", "Home Improvement", "Business", "Personal Loan",
            "Student Use", "Auto", "Baby & Adoption", "Boat", "Cosmetic Procedure",
            "Engagement Ring", "Green Loans", "Household Expenses", "Large Purchases",
            "Medical/Dental", "Motorcycle", "RV", "Taxes", "Vacation", "Wedding",
            "Other", "Not Applicable")

pld$ListingCategory <- factor(pld$ListingCategory..numeric.,
                              levels = c(1:6, 8:20, 7, 0),labels = labels)
```

```{r}
summary(pld$ListingCategory)
```

Let's take a look at the summary of the _LoanStatus_ variable to see what 
grouping is possible for factor plots:

```{r, echo = TRUE,message=FALSE, warning=FALSE}
summary(pld$LoanStatus)
```

Too many variables means there will be too many colours to make an effective 
visualization. There are 6 'Past Due' factors based on how late the borrowers 
are with their payments and other delinquent customers, it will be much more 
useful visually if they're grouped as one. Let's write the code to do that:

```{r, echo = TRUE,message=FALSE, warning=FALSE}
pld <- pld %>%
  mutate(LoanStatusGroup = ifelse(LoanStatus %in% 
                                  c("Cancelled", "Chargedoff", "Defaulted",
                                    "Past Due (1-15 days)","Past Due (16-30 days)",
                                    "Past Due (31-60 days)","Past Due (61-90 days)",
                                    "Past Due (91-120 days)","Past Due (>120 days)"
                                    ), 0,
                           ifelse(LoanStatus %in%
                                  c("Current", "FinalPaymentInProgress"),1,2)))

pld$LoanStatusGroup <- factor(pld$LoanStatusGroup, levels = 0:2,
                                labels = c("Delinquent","Current","Completed"))
```

Lets create a new variable to indicate the year of Loan Creation
```{r, echo = TRUE,message=FALSE, warning=FALSE}
# Create new variable 
pld$LoanCreationYear <- format(as.Date(pld$ListingCreationDate),"%Y")
```

```{r, echo = TRUE}
# Convert dates to date class using lubridate's ymd_hms() function

x <- as.character(pld$LoanOriginationDate)
pld$LoanOriginationDate <- ymd_hms(x)

# Convert LoanOriginationQuarter to begin with the year using tidyr
# This also makes sure that any plot axis will be in increasing order of year

pld$LoanOriginationQuarter <- as.character(pld$LoanOriginationQuarter)
pld <- pld %>%
         separate (col = LoanOriginationQuarter,
                   into = c("Quarters", "Year"), sep = " ") %>%
         unite(col = LoanOriginationQuarter, Year, Quarters, sep = " ")

pld$LoanOriginationQuarterF <- factor(pld$LoanOriginationQuarter)
```

# Univariate Plots Section
Here we perform some preliminary exploration of the dataset by running some 
summaries of the data and create univariate plots to understand the structure of
the individual variables in the dataset


### Loan Characteristics

Since I'm dealing with loan data, the first thing I want to check is the loan 
amounts being requested. Let's make a simple histogram showing just that:

```{r echo=FALSE, fig.width=10, message=FALSE, warning=FALSE}
qplot(x = LoanOriginalAmount, data = pld,binwidth = 1000,colour = I("#424242"),
      fill = I("#0077C0"),main = "Histogram of LoanOriginalAmount") +
  scale_x_continuous(breaks = seq(0, 35000, 5000))
```
Summary Stats for LoanOriginalAmount
```{r echo=FALSE,message=FALSE, warning=FALSE}
summary(pld$LoanOriginalAmount)
```

After adjusting the bin width and x-axis breaks, we get a nice histogram showing 
the loans borrowed ranging from $1000 to the maximum amount of $35,000. The bulk of 
the count lies below $15,000, which indicates that most of the borrowers using 
Prosper are looking for small(er) loans. The other thing I noticed was that there
are huge spikes in the count on nice, round values, like $10,000, $15,000, and 
even $20,000 and $25,000. This makes sense; people are likelier to fixate on values
that are easy to remember, like those that are factors of 5 and 10, rather than 
arbitrary values in between.

The next question that needs to be asked is **_Loan Tenure_**. 
Let's answer this question with a histogram
```{r fig.width = 10}
pld %>% 
    ggplot(aes(x = Term / 12)) + geom_histogram(binwidth = 0.5,fill = I("#0077C0")) +
    xlab('Term in Years') +  ylab('Count') + 
    scale_x_continuous(breaks = seq(1, 5, 2)) +  
    stat_count(aes(y=..count.., label=..count..), geom="text", vjust=-.5) +
    theme(plot.title = element_text(face = 'bold',size=10))
```

**Occupation of Borrowers**

Next I want to visualize more data on these borrowers, and there are a variety of
variables that can paint a clearer picture. Let's first see what they do:

```{r, fig.width = 10, fig.height = 9}
ggplot(aes(x = reorder(Occupation, Occupation,function(x) -length(x)),
           fill = I("#0077C0")),data = pld) + xlab("Occupation") +
  geom_bar() +  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```
I decided to opt for a vertical plot, since there are 68 different occupations listed.

The first thing I noticed in this histogram were the two bars a world away from 
the rest of the occupations - and a quick check told me that it was the count for
**'Other'** and **'Professional'**. It seems like the people signing up to get
loans from *Prosper* are not quite willing to provide their job information, and
are opting instead for these two ambiguous options. Most of the professions have
pretty low counts, but the diversity is quite interesting. Among all of them,
relatively popular ones include analysts, accountants, computer programmers,
teachers and executives. Prosper has definitely enabled a wide range of working
people to either borrow or invest money.

**Purpose of Loan** <br>
 Now we will see for what purpose borrowers typically take the loans.

```{r, fig.width = 10}
qplot(x = ListingCategory, data = pld, fill = I("#0077C0")) + 
  scale_y_continuous(limits = c(0, 60000), breaks = seq(0, 60000, 10000)) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```
_Debt Consolidation_ seems by far the most popular choice, with the rest of the
non-ambiguous (or 'Not Applicable') occupations much below the 10,000 mark. 
Acquiring more debt to repay existing debts seems to be the major intent of 
money borrowers using Prosper. Perhaps they want to structure and streamline 
their debt payments to one source instead of dealing with multiple loans from 
varied sources.

**Region (State) of Borrowers** <br>
Where do the borrowers hail from?
```{r, fig.width = 10, fig.height = 9}
ggplot(aes(x = reorder(BorrowerState, BorrowerState,function(x) length(x)),
           fill = I("#0077C0")),data = pld) + 
  geom_bar() +  xlab("Borrower State") +  coord_flip()
```

We see that the states of California, Texas, New York are the top 3 states having
the most no. of borrowers using Prosper Loans. It is not surprising to see 
California heading the list, as it was here that the company was first founded.

```{r, fig.width = 10, US_Map_of_Borrowers}

# Map the 2 letter abbrievations to their full name for map plot
pld$BorrowerStateFullName <- tolower(state.name[match(pld$BorrowerState, state.abb)])

us <- map_data("state")
states <- data.frame(state.center, state.abb)

# Group by stateName 
grp_by_state <- pld %>% 
  group_by(BorrowerStateFullName, BorrowerState) %>%
  summarise(count = n())

# Plot US map with delinquent borrower data
ggplot() + 
  geom_map(data = us, map = us, aes(map_id = region),fill="#ffffff", 
           color="#ffffff", size=0.15) +
  geom_map(data = na.omit(grp_by_state), map=us,color = "#ffffff", size = 0.15,
            aes(fill = count, map_id = BorrowerStateFullName)) +
  geom_text(data = states,aes(x = x, y = y, label = state.abb, group = NULL), size = 2) +
  scale_fill_continuous(low='lightgoldenrod1', high='orangered1', guide='colorbar') +
  labs(x="Borrowers by State", y=NULL) + coord_map("albers", lat0 = 39, lat1 = 45) +
  ggtitle("Geographic Distribution of Prosper Borrowers by State") +
  theme(plot.title = element_text(face = 'bold',colour = 'black', size=11),
        panel.border = element_blank(),panel.background = element_blank(),
        axis.ticks = element_blank(), axis.text = element_blank())
```

Next let's look at the **income range** of these borrowers
```{r, fig.width = 10}
qplot(x = IncomeRange, data = pld, fill = I("#0077C0")) +
  stat_count(aes(y=..count.., label=..count..), geom="text", vjust=-.5,size=3) +
  scale_x_discrete(limits = c("Not employed", "$0", "$1-24,999", "$25,000-49,999", "$50,000-74,999", "$75,000-99,999", "$100,000+", "Not displayed"))
```
 Most borrowers seem to be in the income bracket from 25,000 $ to 75,000 $.
 
**EstimatedEffectiveYield - A better measure for a Lender's Success**

```{r, fig.width = 10}
ggplot(aes(x = EstimatedEffectiveYield), data = pld) +
    geom_histogram(aes(y = ..density..), binwidth = 0.01, na.rm = T, 
                   color = 'darkblue', fill = 'lightblue') +
    theme_hc() +
    scale_x_continuous(limits = c(-0.1, 0.4),breaks = seq(-0.1, 0.4, 0.05)) +
    geom_density(alpha=.2, fill="#FF6666", na.rm = T) + 
    geom_vline(aes(xintercept=mean(EstimatedEffectiveYield, na.rm=T)), 
               color="blue", linetype="dashed", size=1) +
    ggtitle("Distribution of EstimatedEffectiveYield",subtitle = "Mean in dashed Blue") +
    theme(plot.title = element_text(face = 'bold',colour = 'black', size=10),
          plot.subtitle = element_text(face = 'italic',colour = 'darkblue', size=08))
```
Summary Stats for Estimated Effective Yield
```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(pld$EstimatedEffectiveYield)
```
**EstimatedEffectiveYield** is said to be better estimate for the lenders than 
the interest rate because the interest includes _processing fees_, 
_uncollected interest due to borrower being chargedoff_. Plus it also doesn't 
include _late fines_. Hence EstimatedEffectiveYield takes into account all these
things and it is thus a better measure. In the figure above we are trying to see
the distribution of the EstimatedEffectiveYield and we can see that it is multimodal.
We see the  most popular EstimatedEffectiveYield is around 0.3 while the mean is
around 0.17 represented by the blue dotted line. 
  The multimodal pattern shows that there are multiple EstimatedEffectiveYield 
that is popular. Strangely we can also see that the some customers have negative
EstimatedEffectiveYield. This may mean a lot of things. This may mean that their
BorrowerRate is a lot lower than their _service fee rate_ or these customer's 
_uncollected interest on chargeoff_ is lot more or they just never payed the 
late fee and payed back the loans along with the interest always on time.


**Distribution for some continuous/categorical features**

Lets explore the histogram distribution of other features in this dataset, so we
can evaluate them later if they are interesting

```{r, fig.width = 10}
prosperScoreCountPlot <- pld %>%
    filter(!is.na(ProsperScore)) %>%
    group_by(ProsperScore) %>%
    summarise(n = n()) %>%
    ggplot(aes(x = ProsperScore, y = n)) +
    geom_bar(stat = 'identity', position="dodge",fill='#05DBF1') +
    theme_pander() +
    ylab("No. of borrowers") +
    scale_y_continuous(breaks = seq(0, 15000, 2500)) +
    scale_x_continuous(breaks = seq(1, 11, 1))

EmploymentStatusDurationHist <- ggplot(aes(x = EmploymentStatusDuration),data = pld) +
    geom_histogram(binwidth = 10, color = 'white',fill='#05DBF1') +
    scale_x_continuous(breaks = seq(0, 400, 100), limits = c(0, 400)) +
    theme_pander() +
    xlab("Borrower Experience")

prosperRatingCount <- pld %>%
    group_by(ProsperRating..Alpha.) %>%
    summarise(n = n()) %>%
    ggplot(aes(x = ProsperRating..Alpha., y = n)) +
    geom_bar(stat = 'identity', position="dodge",fill='#05DBF1') +
    theme_pander() +
    ylab("\nNo. of borrowers") +
    xlab("Borrower Rating")

StatedMonthlyIncomeHist <- ggplot(aes(x = StatedMonthlyIncome),data = pld) +
    geom_histogram(binwidth = 1000, color = 'white',fill='#05DBF1') +
    scale_x_continuous(limits = c(0, 20000)) +
    theme_pander()

MonthlyLoanPaymentHist <- ggplot(aes(x = MonthlyLoanPayment),data = pld) +
    geom_histogram(binwidth = 50, color = 'white',fill='#05DBF1') +
    scale_x_continuous(breaks = seq(0, 1000, 300), limits = c(0, 1000)) +
    theme_pander()

LoanOriginalAmountHist <- ggplot(aes(x = LoanOriginalAmount),data = pld) +
    geom_histogram(binwidth = 1000, color = 'white',fill='#05DBF1') +
    scale_x_continuous(breaks = seq(0, 25000, 7000), limits = c(0, 25000)) +
    theme_pander()


BorrowerAPRHist <- ggplot(aes(x = BorrowerAPR),data = pld) +
    geom_histogram(binwidth = 0.01, color = 'white',fill='#05DBF1') +
    scale_x_continuous(breaks = seq(0.07, 0.3, 0.05), limits = c(0.07, 0.3)) +
    theme_pander()

BorrowerRateHist <- ggplot(aes(x = BorrowerRate),data = pld) +
    geom_histogram(binwidth = 0.01, color = 'white',fill='#05DBF1') +
    scale_x_continuous(breaks = seq(0.07, 0.3, 0.05), limits = c(0.07, 0.3)) +
    theme_pander()

DebtToIncomeRatioHist <- ggplot(aes(x = DebtToIncomeRatio),data = pld) +
    geom_histogram(binwidth = 0.05, color = 'white',fill='#05DBF1') +
    scale_x_continuous(limits = c(0.0, 0.9),breaks = seq(0.0, 0.9, 0.1)) +
    theme_pander()

grid.arrange(prosperScoreCountPlot,
             EmploymentStatusDurationHist,
             prosperRatingCount,
             StatedMonthlyIncomeHist,
             MonthlyLoanPaymentHist,
             LoanOriginalAmountHist,
             BorrowerAPRHist,
             BorrowerRateHist,
             DebtToIncomeRatioHist,
             ncol = 3, nrow = 3,
             top = "Distribution of Feature variables\n")
```


## Univariate Analysis
We have performed various operations like tidying up the data and adjusting the 
time series to have quarter yearly durations for future plots. We have 
investigated several features of the dataset which will be key to further analysis 
in this EDA.

# Bivariate Plots Section
 In this section we investigate relationships between variables, both the
 main features as well as the supporting variables.

**Do Lenders prefer borrowers with better Prosper Score ?**

Now lets see what is the distribution of EstimatedEffectiveYield depending on the
different [**ProsperScore**](https://www.prosper.com/plp/general-prosper_score/).
This is important because we want to answer the question: DO LENDERS GET MORE 
EstimatedEffectiveYield IF THEY HAVE BETTER ProsperScore ?

We are using violin plot instead of box plot for this.

```{r, fig.width = 10}
pld$ProsperScore <- factor(pld$ProsperScore)
ggplot(aes(x = ProsperScore, y = EstimatedEffectiveYield, fill=ProsperScore), 
       data = subset(pld, !is.na(pld$ProsperScore) & !is.na(pld$EstimatedEffectiveYield))) +
    geom_violin(trim = F, scale = "width") +
    stat_summary(fun.y=median, geom="point", size=2, color="black") +
    scale_fill_manual(values=colorRampPalette(c("lightgreen", "#05DBFF"))(11)) + 
    theme_minimal() +
    ylab('Effective Yield of Lenders') +
    ggtitle("Effective Yield by Prosper Score",subtitle = "Indicator for Lenders preference") +
    theme(plot.title = element_text(face = 'bold.italic',colour = 'black',size=12),
          plot.subtitle = element_text(face = 'bold',colour = 'darkgreen',size=09)) +
    guides(fill = F)
```

We see some interesting trends here. Here a higher Prosper score means better 
prospects (with 11 being the best, or lowest risk, score) and lower Prosper score
means poor prospects for the lenders (worst, or highest risk, score, is a 1). 
We can see that for lower ProsperScore, distribution of effective yield is a lot 
more than the higher ProsperScore. This may mean that lenders charge a variety 
of interest rate for the borrower with poor prospects as compared to borrowers
with better prospect. We can also notice how the median (represented by the 
black dot)is decreasing as ProsperScore is increasing. This may mean that lenders
give more relaxations to borrowers with better ratings as compared to borrowers
with poor rating. 

_Does that mean lenders trust and like borrowers with better ProsperScore_? <br>
  A higher EstimatedEffectiveYield for lesser ProsperScore borrowers may be due
to high late fines because lesser ProsperScore borrowers are more prone to fail 
to repay their loans in a timely fashion every month. A little more analysis may
reveal it more. The reason we need more exploration on this is because 
EstimatedEffectiveYield includes more things such as late fines and doesn't 
include processing fees and others. So, let's see if borrower's interest rate 
shows the same trend for each ProsperScore categories or not because interest 
rates doesn't include late fines. 

```{r, fig.height=5, fig.width=10}

borrower_apr <- ggplot(aes(x = ProsperScore,y = BorrowerAPR),
                       data = subset(pld, !is.na(pld$ProsperScore) & !is.na(pld$BorrowerAPR))) + geom_boxplot(aes(group = cut_width(ProsperScore, 1)),outlier.color = I("#FBB660"),outlier.shape = 1,outlier.alpha = 0.5) + theme_pander()

borrower_rate <- ggplot(aes(x = ProsperScore,y = BorrowerRate),
                        data = subset(pld,!is.na(pld$ProsperScore) & !is.na(pld$BorrowerRate))) + geom_boxplot(aes(group = cut_width(ProsperScore, 1)),outlier.color = I("#FBB520"), outlier.shape = 1,outlier.alpha = 0.5) +  theme_pander()

grid.arrange(borrower_apr, borrower_rate,nrow = 1, ncol = 2,
             top = "Interest Rate Distributions")
```

Finally things are revealed much better now! We can clearly observe that for both
_BorrowerAPR_ and _BorrowerRate_ which are metrics for interest rates, we see a 
declining trend as the _ProsperScore_ increases. This justifies even more the 
fact that lenders somehow prefer to charge less for the borrowers with better
ProsperScore as compared to borrowers with inferior ProsperScore.

**Listing Category Vs. Loan Amount and Interest Rate**

```{r, fig.width = 10}

ggplot(pld, aes(ListingCategory,LoanOriginalAmount,group = ListingCategory,
                fill = I("#F5F5F4"), color = I("#506E85"))) +
  geom_boxplot(outlier.color = I("#FBB448"), outlier.shape = 1) +
  theme_minimal() + ylab("Loan Amount in USD") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

ggplot(pld, aes(ListingCategory,BorrowerRate, group = ListingCategory,
                fill = I("#F5F5F4"), color = I("#506E87"))) +
  geom_boxplot(outlier.color = I("#FBB448"), outlier.shape = 1) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
# grid.arrange(lc_loa,lc_br,ncol = 1, nrow = 2)
```

**Summary Statistics of Loan Original Amount for each Listing Category**

```{r}
by(pld$LoanOriginalAmount,pld$ListingCategory,summary)
```

```{r eval=FALSE, message=FALSE, warning=FALSE, include=FALSE}
**Summary Statistics of Borrower Rate for each Listing Category**
by(pld$BorrowerRate, pld$ListingCategory, summary)
```

In the above two plots, I examine the loan category, loan amounts and borrowing 
rate. Debt consolidation as expected has the highest loan amounts across 
categories; however, unexpectedly "Baby and Adoption" category is also the highest
(tied) loan amount across categories. It would be interesting to investigate the
total number of borrowers across these two categories, one would expect debt 
consolidation to have a large distribution of borrowers.

Also of note, the "Student Use" category loan amount it quite low. Given the well
documented high levels of student debt, we can assume students didn't suddenly 
stop having debt but instead don't use Prosper to manage their debt or this 
category covers items other than student tutition such as lunches, dinners and books.

The borrower's interest rate doesn't seem to have any particular surprises. The 
mean borrowing rate across all categories tend to vary between 0.15 and 0.25. 


**Distribution of Loans over the Years**

```{r, fig.width = 10}
# Investigate distribution of loan creation year
ggplot(pld, aes(pld$LoanCreationYear,fill = factor(LoanStatusGroup))) + 
  geom_bar() + xlab("Year") + ylab("Loans Created") +
  theme(legend.position="top") +
  scale_fill_manual(name = "Loan Status",values = c("#FA4659","#FBB448","#2EB872"))
```
In this plot we see basically see the growth story of _Prosper_ over the years
since its inception. This seems to be a very interesting plot that is worthy of
further investigation.

**Debt-to-Income Ratio and Monthly Income**

```{r, echo = FALSE}
loans.status <- pld %>%
  filter(LoanStatusGroup %in% c("Completed", "Delinquent")) %>%
  select(StatedMonthlyIncome,DebtToIncomeRatio,LoanStatusGroup)
```

```{r, fig.width = 10}
# monthly income vs debt to income ratio, highlighting delinquent borrowers
ggplot(loans.status, aes(StatedMonthlyIncome,DebtToIncomeRatio, 
                         colour = factor(LoanStatusGroup))) +
  geom_point(alpha = 0.25) +  theme_minimal() +
  scale_x_continuous(limits = c(0,20000)) +
  scale_y_continuous(limits = c(0, 1)) +
  theme(legend.title=element_blank()) +
  scale_colour_manual(values=c("#FA4659","#2EB872")) +
  geom_density2d()
```

In comparing debt-to-income ratio with a borrower's stated monthly income one 
might expect to see a somewhat obvious, trend that delinquent borrowers would have
a lower monthly income and a higher debt-to-income ratio. With overplotting in the
scatterplot two additional techniques were used to more clearly reveal any trends
or unexpected results. Borrowers with stated incomes over $20k and debt-to-income
ratios over 1 were considered as outliers and removed from the plot. 

The density contour lines show a high concentration of delinquent borrowers earn
less than $2500 a month but have a low debt-to-income ratio of under 0.50 (or 50%).
The plot also suggests a negative correlation between monthly income and 
debt-to-income ratio,i.e the more a borrower makes in monthly income the lower 
their debt-to-income ratio; However, this does not guarantee the loan will not 
go into delinquency.

After further review from [online resources](https://smartasset.com/credit-cards/what-is-a-good-debt-to-income-ratio),
it seems that a typical "good" debt-to-income ratio is under approx. 36%.

## Bivariate Analysis
To identify significant relationships between the different variables in the
dataset lets go for a correlation matrix.

**Let's select only the variables of interest.**

```{r echo=TRUE}
colsToKeep <- c("Term","EstimatedLoss","EstimatedEffectiveYield","LenderYield",
               "CreditScoreRangeLower","CreditScoreRangeUpper","MonthlyLoanPayment",
               "LoanOriginalAmount","TotalProsperLoans","Investors","BorrowerAPR",
               "BorrowerRate","DebtToIncomeRatio","StatedMonthlyIncome"  )

# Keep only variables in colsToKeep (columns to keep)
prosperloans <- subset(pld, select=colsToKeep)
```
A correlation matrix will be used to calculate the Pearson coefficients of the
various features in the dataset in order to help start the investigation process
where by the variables with strong and/or weak relationships will be reviewed.

**Correlation Matrix**

```{r, fig.width = 10, Correlation_Matrix}
# Pearson correlation coefficients, using pairwise observations (default method)
# Non-numeric columns automatically removed/ignored
ggcorr(prosperloans, label = TRUE, label_size = 3, hjust = 0.8, size = 2.5, 
       color = "black", layout.exp = 2)
```
There are lots of correlated variables  in the data set. The most correlated 
variables (corr = 0.911) are _Loan origination amount_ and _monthly loan payment_, 
which is not surprising.

Borrower rate is negatively correlated with Prosper score and credit score.
Debt income ratio is slighlty correlated with credit score.

Prosper score is also correlated with loan origination amount and monthly payment.

# Multivariate Plots Section

In this section, we will further examine the relationships between multiple 
variables of the dataset. We will continue with the strong relationships identified
from the correlation matrix in the Bivariate Plots section.

**Estimated Effective Yield by Estimated Loss**
```{r fig.width=10, echo=FALSE}
eyel_term <- ggplot(data=pld,aes(EstimatedEffectiveYield,EstimatedLoss)) +
  geom_jitter(na.rm = TRUE,alpha=1/2,aes(color=factor(Term)))

eyel_ps <- ggplot(data=pld,aes(EstimatedEffectiveYield,EstimatedLoss)) +
  geom_jitter(na.rm = TRUE,alpha=1/2,aes(color=factor(ProsperScore)))

grid.arrange(eyel_term,eyel_ps,nrow = 1, ncol = 2,
             top = "EstimatedEffectiveYield Vs. EstimatedLoss")
```
The correlation matrix showed a high positive correlation (0.9) between 
_Estimated Effective Yield_ and _Estimated Loss_. Investigating this relationship
further constrained on Term of the Loan and ProsperScore reveals some interesting insights.
  The short term 1 year loan, seems to be having a higher exponentially increasing
Estimated loss in comparison to the other term loans. The widely popular 3 year 
loan shows Estimated Losses all over the place.

The plot on the right factored by Prosper Score indicates a logical and 
expected outcome, lower Prosper scores (close to 1) indicating high risks show
a higher Estimated Loss and for high Prosper Scores indicating very low risks
the Estimated Loss is very low.


**Monthly payment and loan amount**
```{r fig.width=10, echo=FALSE}
ggplot(data=pld,aes(LoanOriginalAmount,MonthlyLoanPayment)) +
  geom_jitter(aes(color=factor(Term)),alpha=1/3) + 
  geom_smooth(method = "lm")
```
There are three clusters of points: one is the 1 year term loan, another is the
3 year term period loan, and the last cluster is the 5 year term period loan. A
linear regression line has also been added to the plot, which we will explore at
the end of this section on Multivariate Plots.

We can try to explain the variance.
```{r fig.width=10,  echo=FALSE}
# ProsperScore was interpreted as factor. Lets convert it onto numeric.
pld$ProsperScore=as.numeric(pld$ProsperScore)

ggplot(data=subset(pld,!is.na(ProsperScore) & Term=='36'),
       aes(LoanOriginalAmount,MonthlyLoanPayment))+
  geom_point(aes(color=ProsperScore),alpha=1/5) + 
  scale_color_continuous(high="lightgreen",low="darkblue")
```

We observe the variance explained by risk, represented by the ProsperScore.
The bottom of the scatter plot is dominated by loans with a ProsoperScore equal
to 11, which represents loans with low risks. The top of the scatter plot is 
dominated by loans with a ProsoperScore equal to 4, which represents loans with
higher risks. Loans with a loan amount higher than $25000 are mostly dominated by
a ProsperScore equal or superior to 1.


## Multivariate Analysis

We have observed that the Prosper score has an influence on montly payments for 
the same loan origination amount and same term. This Score should be a way to 
quantify the probability that the loan is going to default or not. The Prosper
Score also revealed the quantum of Estimated Loss that is likely to be incurred 
by the lender. However when we look at the Prosper Score alone it doesn't seem 
be a great indicator for the investor if he should invest or not, for the 
features considered in the Multivariate Analysis.

#### OPTIONAL:

**Linear Mixed Effects Model of Monthly Loan Payment vs. Loan Original Amount**:
```{r echo=FALSE}
fit1 <- lm(MonthlyLoanPayment ~ LoanOriginalAmount, data = pld)
summary(fit1)
```
_LoanOriginalAmount_ should be the most important factor to decide what would be 
the MonthlyLoanPayment. As we can see that the correlation has very good Adjusted
R-squared value of 0.845 or almost 85%. The Residual Standard Error is however 
quite high almost 70 which indicates the average performance of the linear model.

# Final Plots and Summary
 In this section we select and refine three plots from all the above explorations
 and present here a summary of some of the most interesting findings.

### Plot One
```{r, echo = FALSE}
loans.defaults <- pld %>% 
  group_by(Quarter = LoanOriginationQuarter, LoanStatusGroup) %>%
  summarise(Loans = n()) %>%
  arrange(Quarter, LoanStatusGroup)
```

```{r eval=FALSE, fig.width=10, message=FALSE, warning=FALSE, include=FALSE}
# ggplot(aes(x = Quarter, y = Loans, fill = LoanStatusGroup),data = loans.defaults) +
#   geom_bar(stat = "identity")  + ylab("Loans Created") +
#   theme(axis.text.x = element_text(angle = 90, vjust = 0.5),legend.position="top") +
#   scale_fill_manual(name = "Loan Status",values = c("#FA4659","#FBB448","#2EB872")) +
#   ggtitle("Status and Growth of Prosper Loans") +
#   theme(plot.title = element_text(face = 'bold',colour = 'black',size=11))
```
It is very worthwhile to investigate the creation date distribution of loans 
in the dataset. This is important as this could cause bias in further
investigation, for example, if the majority of the data was observed during 
2008-2009 [(the financial crisis)](https://en.wikipedia.org/wiki/Financial_crisis)
this could skew the data towards having a majority of delinquency.
```{r, fig.width = 10}
f <- list(size = 12,color = "#7f7f7f")
x <- list(titlefont = f,tickangle = 90)
y <- list(title = "No. of Loans",  titlefont = f)
title = "Status and Growth of Prosper Loans"
plot_ly(data = loans.defaults, x = ~Quarter, y = ~Loans,type = 'bar',
        color = ~LoanStatusGroup,colors = c("#FA4659", "orange", "#2EB872"))%>%
  layout(xaxis = x, yaxis = y, title = title, margin = list(b = 60),
         legend = list(x = 0.01, y = 0.95),barmode = 'stack')
```

#### Description One
  In the above plot we further investigate the plot we made in the Bivariate section.
We have expanded the X-axis to include Yearly quarters instead of just showing the year.
Here we see some dampening in the business of _Prosper Funding LLC_ which can be
attributed to the [financial crisis](https://en.wikipedia.org/wiki/Financial_crisis_of_2007%E2%80%932008) 
of 2008. There was a lot of delinquent (_bad performing_) loans. Post 2009, the 
company has restructured its business models and policies and has been growing
nicely since then.

### Plot Two
```{r fig.width=10, echo=FALSE}
f <- list(size = 12,color = "#7f7f7f")
x <- list(title = "Loan Origination Amount ($)" ,  titlefont = f)
y <- list(title = "Monthly Loan Payment ($)",  titlefont = f)
df<- subset(pld,!is.na(ProsperScore) & Term=='36')
title = "Monthly Loan Payment by Loan Origination Amount for 3 year loan"
plot_ly(data = df, x = ~LoanOriginalAmount, y = ~MonthlyLoanPayment,
        mode = "markers",color = ~ProsperScore)%>%
  layout(xaxis = x, yaxis = y, title = title)
```

#### Description Two
 Here we select the plot of **Loan Origination Amount Vs. Monthly Loan Payment**
from the Multivariate Plots section but restrict the data for 3 year Term loans,
as it is the most popular term for borrowers.

### Plot Three

```{r echo=FALSE, fig.width=10}
borrower_apr <- ggplot(aes(x = ProsperScore,y = BorrowerAPR,color=ProsperScore,
                           fill = I("#F5F5F4"), color = I("#506E85")), 
       data = subset(pld, !is.na(pld$ProsperScore) & !is.na(pld$BorrowerAPR))) +
    geom_jitter(shape=2, position=position_jitter(0.2), col = "deepskyblue",
                alpha = 1/12) +
    geom_boxplot(aes(group = cut_width(ProsperScore, 1)),
                 outlier.shape=NA, col = "dodgerblue2") +
    stat_summary(fun.y=median, geom="point", size=2, color="black") +
    theme_pander() +
    xlab('\n Prosper Score') +
    ylab('Annual Percentage Rate') +
    ggtitle("Borrower Annual Rate", subtitle = "for each borrower group") +
    theme(plot.title = element_text(face = 'bold',colour = 'dodgerblue2',size=12),
          plot.subtitle = element_text(colour = 'black',size=10))

borrower_rate <- ggplot(aes(x = ProsperScore,y = BorrowerRate,color=ProsperScore,
                            fill = I("#F5F5F4"), color = I("#506E85")), 
       data = subset(pld, !is.na(pld$ProsperScore) & !is.na(pld$BorrowerRate))) +
    geom_jitter(shape=2, position=position_jitter(0.2), col = "indianred2",
                alpha = 1/12) +
    geom_boxplot(aes(group = cut_width(ProsperScore, 1)),
                 outlier.shape=NA, col = "indianred4") +
    stat_summary(fun.y=median, geom="point", size=2, color="black") +
    theme_pander() +
    xlab('\n Prosper Score') +
    ylab('Borrower Interest Rate') +
    ggtitle("Borrower Interest Rate", subtitle = "for each borrower group") +
    theme(plot.title = element_text(face = 'bold',colour = 'indianred2',size=12),
          plot.subtitle = element_text(colour = 'black',size=10))

grid.arrange(borrower_apr,borrower_rate, nrow = 2, ncol = 1)
```

```{r eval=FALSE, fig.width=10, warning=FALSE, include=FALSE}
pld$ProsperScore=as.numeric(pld$ProsperScore)
f <- list(size = 12,color = "#7f7f7f")
x1 <- list(title = "ProsperScore by Risk Factor" ,  titlefont = f)
y1 <- list(title = "Annual Percentage Rate",  titlefont = f)
y2 <- list(title = "Borrower Interest Rate",  titlefont = f)
title1 = "Borrower Annual Rate by ProsperScore"
title2 = "Borrower Interest Rate by ProsperScore"
df1 = subset(pld, !is.na(pld$ProsperScore) & !is.na(pld$BorrowerAPR))
df2 = subset(pld, !is.na(pld$ProsperScore) & !is.na(pld$BorrowerRate))
p1 <- plot_ly(data = df1, x = ~ProsperScore, y = ~BorrowerAPR,
             color = ~ProsperScore, type = "box") %>%
  layout(xaxis = x1, yaxis = y1, title = title1)
p2 <- plot_ly(data = df2, x = ~ProsperScore, y = ~BorrowerRate,
             color = ~ProsperScore, type = "box") %>%
  layout(xaxis = x1, yaxis = y2, title = title2)
subplot(p1,p2,nrows = 2,shareX = TRUE)
```

#### Description Three
_Do lenders prefer borrowers with better ProsperScore ?_ <br>
This plots shows that interest rates are affected by the Prosper Score for 
varying risk factors. As the Prosper score improves, the interest rate shows a 
declining trend visible from the black points.
------

# Reflection
The prospser loans dataset contains over 100k observations with 81 variables 
spanning across 9 years. Understanding the variables, terminology and general 
domain knowledge of financial peer-to-peer lending was the first obstacle in 
approaching this dataset.But the overall experience has been quite rewarding in
this EDA. I had learned many tools and concepts in Data Analytics with R in this process.

We have been able to analyze that Prosper's portfolio has really changed between
2009 and 2014. It seems that Prosper doesn't only use credit scores to assess risk
but also several other variables that could be in the dataset, in addition to 
external data not included in the dataset available.

Instead of just using the Prosper score to decide whether or not to invest, it 
would be interesting to build a predictive model that quantifies if the person is
going to default on their loan. We could use, for example, a logistic regression.
However there are lots of variables that are correlated, such as loan origination
amount, credit score, and Prosper score. It's important to carefully choose which
variables should be included in the model.

Additional data would also enhance this dataset. Having the borrower's age and 
sex would allow analysis to possibly discover trends among men and women or 
young and old. Also, population and state-average-income features, would allow
for discovery of the type of environment the borrower lived in. 

## References
* https://en.wikipedia.org/wiki/Charge-off
* https://en.wikipedia.org/wiki/Default_(finance)
* https://www.prosper.com/help/topics/how-to-read-a-loan-listing
* http://www.investopedia.com/terms/d/delinquent.asp
* http://docs.ggplot2.org
* http://zevross.com/blog/2014/08/04/beautiful-plotting-in-r-a-ggplot2-cheatsheet-3
* https://en.wikipedia.org/wiki/List_of_states_and_territories_of_the_United_States
* http://stackoverflow.com/a/29618201/1343001
* http://www.investopedia.com/ask/answers/100314/what-difference-between-interest-rate-and-annual-percentage-rate-apr.asp
* https://briatte.github.io/ggcorr